### 2023-02-13 article.tex
Use '\cite' with or without braces

The concept of paradata (although not named as such yet) 
was introduced by Couper and colleagues,
who developed a computer-assisted interview program
that, among others, records all key strokes,
measures the time used to answer each question 
and even the time the monitor is turned off \cite{couper1998measuring}.
The goal in that context was to assess and improve survey quality.

@@

This paper defines paradata as 'data about the data collection 
process' \cite{choumert2019using}.
This definition is chosen for its simplicity, 
yet there are multiple related and/or more nuanced 
definitions of paradata (see for example \cite{huvila2022improving} 
and \cite{skold2022interrogating}).

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful way, 
the documentation and data should have co-evolved, 
and the paradata should be computer-friendly \cite{huvila2022improving}.
At first glance, these suggestions appear to work well for code
and will be discussed in detail below.

@@

There are multiple reasons why useful paradata matter. 
Most obvious is that having useful paradata gives an understanding how
data is produced.
This knowledge helps researchers from different fields to understand each other
and collaborate.
Additionally, useful open data is needed for Open Science to convincingly show
its benefits.
Finally, in computational fields, 
it can help understand how scholarly knowledge is produced
\cite{huvila2022improving}.
This paper discusses these general reasons applied to code.

@@

To determine how useful code is, it needs to be available.
However, it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
42\% of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where 60\% of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
For articles in Science magazine, 12\% of 204
studies published the code \cite{stodden2018empirical} 
(note that these were years 2011-2012).
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced \cite{haibe2020importance}.

@@

When code of an academic paper is not published, 
one could contact the corresponding author and request it.
However, the response rate of corresponding authors 
with a request for data in computational fields is around 50\% 
\cite{manca2018non, stodden2018empirical, teunis2015corresponding} 
(the field of emergency medicine seems to be a pleasant outlier, 
with 73\% of 118 emails being replied \cite{o2003email}).
When getting an answer of the corresponding author, 
48\% out of 134 replied emails will actually result 
in a sharing of the code \cite{stodden2018empirical}.
The responses of unwilling authors (see \cite{stodden2018empirical} for 
some real examples) can come across as so caustic, 
that one may be excused from not contacting an a corresponding author.

@@

Genetic epidemiology is a field within biology that
tries to determine the spread of heritable traits 
and their underlying biological mechanism.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east Asia and south Africa \cite{storhaug2017country}.
The trait is caused by the genetic make-up, or 'genotype'
\iffalse
(see Table \ref{tab:definitions} in the Supplementary Materials
for the definitions of terms used in this paper)
\fi
, of a person.
The trait, also called 'phenotype', 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations, when relevant enough, are used to 
create so-called 'gene panels', 
where the location of the gene causing 
an association is measured specifically, 
to detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 \cite{igl2010northern}. 
The approximately 1000 participants were initially mostly surveyed
about lifestyle \cite{igl2010northern} and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes \cite{johansson2013identification},
(2) the phenotypes, in this case, concentrations of certain proteins in the 
blood \cite{enroth2014strong,enroth2015effect}.

@@

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, Ahsan and colleagues show that
SNP \verb|rs12133641| is highly correlated (p-value is $3.0^{-73}$
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
,
for $n$ = 961 individuals) with protein IL6RA \cite{ahsan2017relative}.
The direction of the association is also concluded:
the more guanines are present at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. The amount of variance that can be explained by an association (i.e.
the \begin{math}R^2\end{math}) is rarely 100\%, which means that a trait (in
this case, the concentration of IL6RA) cannot be perfectly explained
from the genotype (in this case, SNP \verb|rs12133641|) alone. 
In this example, as much as 43\% of the variance 
can be attributed to an individuals' genotype
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. Additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day the measurement 
was done), lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the additional variation.

@@

For computational science, it may appear to be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, 
only 18\% of 180 studies are easily reproducible \cite{stodden2018empirical}.
To some, it appears that 
the academic culture to reproduce results 
has been lost over time \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
being the exception.
One suggested way forward is to make the reproduction of 
research a minimal requirement for publication \cite{peng2011reproducible}.

@@

A genetic epidemiologist works with sensitive data as well:
the genetic sequences of participants are private \cite{clayton2019law}.
For research to be reproducible one needs both the code and the data
to reproduce the (hopefully) same results.
This problem is discussed in section \ref{sec:sensitive-data}.

@@

We ignore the choices of words and style of this sentence: what is
important is the content: an unpaired one-tailed T test is performed.
Taking a look at the (in this case, the programming language R \cite{r}) code, 
we find the following line:

@@

Note that there is another side of this same coin:
correctly reporting the t-test in a reproducible ways
is known to be a problem, with only 80 out of a selected 179 papers
fully reporting the parity of t-test (i.e. paired or unpaired) 
\cite{weissgerber2018we}.
In those unresolved 99 papers, 
the code (if published) would clarify this.
\fi

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
As humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
\cite{baggerly2009deriving, vable2021code},
and this example is not too far off from the
examples given in that paper.

@@

Code is rarely preserved \cite{barnes2010publish}.
If it is, keeping it running is even more rare.
This section discusses the preservation of code for a short, medium
and long term.

@@

To preserve code for a short term, 
a code hosting website is a good first step.
A code hosting website is a website where 
its users can create dedicated pages (called 'repositories')
for a project, upload code and interact with that code.
There are multiple code hosting websites, 
with GitHub being the most popular one \cite{cosentino2017systematic}.
For example, the \LaTeX code of this chapter is hosted on GitHub 
as well, at \url{https://github.com/richelbilderbeek/chapter_paradata}.
The use of code hosting websites
has increased strongly \cite{russell2018large},
accommodates collaboration \cite{perez2016ten}
and improves transparency \cite{gorgolewski2016practical},
due to its inherent computer-friendliness.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

\iffalse
An example from the author is the website \cite{bbbqarticleissue157},
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane}, 
as shown in Figure \ref{fig:bbbqarticleissue157}.
\fi

@@

Hosted code commonly keeps a history of file changes,
a so-called commit history,
where a commit is the description of the change, 
as well as the changes in one or more files.
For example, GitHub
uses a version control system called \verb|git| (hence its name).
In practice, this means that when a change is made to the code,
a new version is created. In the case that the change was harmful,
one can go back to an earlier version and continue again from there.
It is a general recommendation to put version control
on all human-produced data \cite{wilson2014best},
as well as openly working on the code from the start \cite{jimenez2017four}.
Half of 56 Science magazine corresponding authors 
have such a version control mechanism \cite{stodden2018empirical}.

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. 
\iffalse
Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE \cite{etienne2020daisie}
have made to its code.
\fi
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

@@

To preserve code for a longer timespan, 
continuous integration ('CI') allows one to 
verify if code still works and, if not, to be notified on the need for an update.
Things change: from operating systems, to libraries that code uses,
to the code itself. 
Embracing those changes is recommended \cite{beck2000extreme} 
and useful code does so.

@@

Some code hosts, among others, GitHub, 
allow the researcher/user to start software scripts upon uploading a change.
This is useful to determine the reproducibility of the code: 
one such script should run the code (on trivial data) 
to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediately see which small change was the cause of this.
Using a CI service (such as GitHub Actions for GitHub), 
is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. 
\iffalse
For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a \LaTeX document (i.e. this paper)
could be built.
\fi

@@

For paradata to be useful, it has to be computer-friendly, yet 
'the best paradata does not necessarily look like 'data' at all for its
human users' \cite{huvila2022improving}.
There are features of code that humans find useful,
without directly being able to measure these.
In the end, code is just 'another kind of data' and should be designed 
as such, for example, by using tools to work on it \cite{wilson2022twelve}.

@@

A first example is to use a tool to enforce a coding style 
(e.g. the Tidyverse style guide \cite{wickham2019advanced} for R,
or PEP 8 \cite{van2001pep} for Python),
as following a consistent coding style improves software quality \cite{fang2001}.

@@

A second example is to use a tool to enforce a low cyclomatic complexity.
The cyclomatic complexity is approximately defined 
as the number of independent paths that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false,
resulting in the body of the \verb|if| statement being either
executed or ignored.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}.

However, for code to work longer, one can put the code and its dependencies in
a container, that should be best in standing the test of time,
as the degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here a more modern (and in this context superior)
definition is used: software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where that paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
this paper recommends a modern solution that potentially has more 
longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

@@

To repeat features of useful paradata: paradata should be comprehensive, 
documented appropriately, documented in co-evolution with the data,
and the data should be friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.
However, code has multiple uses, which have different requirements
for being useful paradata.

Code can be used to reproduce, to replicate and to 
extend a computational experiment \cite{benureau2018re} and
each of these uses has different requirements for the code to be useful paradata.
In this section, the recommendations how to make paradata useful
is applied to code and compared to existing best practices in software 
development, for the intended use of the code.

@@

For code to be ideal paradata, it must be usefully 
documented \cite{huvila2022improving}.
What is 'useful' is determined by its intended use: 
to reproduce, replicate or extend (see section \ref{sec:making-code-useful-paradata}).
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

For purposes of reproducing (same code, same data) code,
it should at least be documented how the code is run.
However, 57\% out of 56 Science papers with obtainable code (in 
total there were 180 papers) have documented their code well enough to run it 
and to be understood what it ought to do \cite{stodden2018empirical}.

However, when an independent researcher can redo an experiment, 
it is possible (even ideal!) that no code is read at all.
Within that context, it could argued that no further documentation is needed.
However, all code in general should be documented 
'adequately' \cite{peng2006reproducible},
ideally writing code in such a way that it becomes 
self-explanatory \cite{wilson2014best}
and for the remaining code to document the reasons behind it, 
its design and its purpose \cite{wilson2014best}.

For purposes of extending a study and its code, 
documentation becomes even more important,
as the code will be read and modified.
The extent of investing time in documenting code is recommended
to be proportional to the intended reuse \cite{pianosi2020successfully}
and there exists a clear relationship between the reuse 
of code and its documentation 
effort \cite{cosentino2017systematic,hata2015characteristics}.

@@

For code to be ideal paradata, 
its creation and documentation need to align, 
\emph{not least because they shape each other} \cite{huvila2022improving}.
This emphasised part of the quote resonates strongly with the idea of 
literate programming \cite{knuth1984literate},
in which documentation and code is developed hand-in-hand.
Literate programming is the practice of writing a document mixed with code that 
will be run.
Contemporary examples of this idea are, among others, 
vignettes \cite{wickham2015r} (for the R programming language \cite{r}) 
and Jupyter notebooks \cite{wang2020assessing} 
(for the Julia \cite{Julia-2017}, Python \cite{van1995python} and R \cite{r}
programming languages).
rOpenSci, a community that, among others, reviews programming
code (\cite{ram2013ropensci,ram2018community}), is 
one example of where literate programming is enforced:
code must have extensive runnable examples
as part of the documentation \cite{ropensci_2021_6619350}.
Developing code and documentation at the same time 
leads to both better code and documentation \cite{reenskaug1989environment}.
In general, when developing software, 
it is recommended to 
to write documentation while writing software,
as well as to include many examples \cite{lee2018ten}.

@@

Code should be distributed in standard ways \cite{peng2006reproducible}
(see section \ref{sec:code-availability}
for the general availability of code).
The standard way to publish code is to use a code hosting website
(see subsection \ref{subsec:code-hosting}).

@@

Code must be more extensive when it is (intended to be) 
used on different data,
as then 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
One example is when doing a t-test:
to compare the means of two distributions, each distribution
must have at least two values
and, if not, the programming language should signal the error clearly.
In genetic epidemiology, it is common to have incomplete or missing data,
so analyses should take into account that there are not always enough
observations to do a t-test.
The analytical code should clearly signal this problem.

@@

Coding errors are extremely common 
and contribute to the reproducibility crisis in science \cite{vable2021code}.
Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic).

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

@@

Having a 100\% code coverage is not uncommon,
as there is a common software development practice that 
strives for this, called test-driven development ('TDD'). 
TDD is a software development strategy in which tests are written
before new 'real' code is written.
TDD improves code quality \cite{alkaoud2018quality,janzen2006test}
and writing comment and documentation mixes well \cite{shmerlin2015document}.

@@

It is considered good practice
to add a software license \cite{jimenez2017four},
so that it is clear that the software can be reused.
About two-thirds of 56 Science magazine corresponding authors 
that shared their code supply a software license \cite{stodden2018empirical}.

@@

In an informal Twitter poll,
more than half of all 315 responding scientists 
have their code rarely or never reviewed \cite{vable2021code},
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
\cite{vable2021code}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best} 
The observation that code is reviewed, makes it likelier to be correct.

@@

For those 22 (out of 180) Science magazine corresponding authors 
that supplied code and that code was estimated to be 
reproducible, one in three had the results reproduced with 
minimal or no effort \cite{stodden2018empirical}.

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data are
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.

>>>>

The concept of paradata (although not named as such yet) 
was introduced by Couper and colleagues,
who developed a computer-assisted interview program
that, among others, records all key strokes,
measures the time used to answer each question 
and even the time the monitor is turned off (\cite{couper1998measuring}).
The goal in that context was to assess and improve survey quality.

@@

This paper defines paradata as 'data about the data collection 
process' (\cite{choumert2019using}).
This definition is chosen for its simplicity, 
yet there are multiple related and/or more nuanced 
definitions of paradata (see for example \cite{huvila2022improving} 
and \cite{skold2022interrogating}).

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful way, 
the documentation and data should have co-evolved, 
and the paradata should be computer-friendly (\cite{huvila2022improving}).
At first glance, these suggestions appear to work well for code
and will be discussed in detail below.

@@

There are multiple reasons why useful paradata matter. 
Most obvious is that having useful paradata gives an understanding how
data is produced.
This knowledge helps researchers from different fields to understand each other
and collaborate.
Additionally, useful open data is needed for Open Science to convincingly show
its benefits.
Finally, in computational fields, 
it can help understand how scholarly knowledge is produced (\cite{huvila2022improving}).
This paper discusses these general reasons applied to code.

@@

To determine how useful code is, it needs to be available.
However, it is not common to publish the code of an experiment or analysis 
(\cite{stodden2011trust,read2015sizing}) (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
42\% of 454 SIGGRAPH papers supply computer code (\cite{bonneel2020code}).
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where 60\% of 62 studies supplied the code 
to redo the analysis (\cite{obels2020analysis}).
For articles in Science magazine, 12\% of 204
studies published the code (\cite{stodden2018empirical}) 
(note that these were years 2011-2012).
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced (\cite{haibe2020importance}).

@@

When code of an academic paper is not published, 
one could contact the corresponding author and request it.
However, the response rate of corresponding authors 
with a request for data in computational fields is around 50\% 
(\cite{manca2018non, stodden2018empirical, teunis2015corresponding}) 
(the field of emergency medicine seems to be a pleasant outlier, 
with 73\% of 118 emails being replied (\cite{o2003email})).
When getting an answer of the corresponding author, 
48\% out of 134 replied emails will actually result 
in a sharing of the code (\cite{stodden2018empirical}).
The responses of unwilling authors (see \cite{stodden2018empirical} for 
some real examples) can come across as so caustic, 
that one may be excused from not contacting an a corresponding author.

@@

Genetic epidemiology is a field within biology that
tries to determine the spread of heritable traits 
and their underlying biological mechanism.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east Asia and south 
Africa (\cite{storhaug2017country}).
The trait is caused by the genetic make-up, or 'genotype'
\iffalse
(see Table \ref{tab:definitions} in the Supplementary Materials
for the definitions of terms used in this paper)
\fi
, of a person.
The trait, also called 'phenotype', 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations, when relevant enough, are used to 
create so-called 'gene panels', 
where the location of the gene causing 
an association is measured specifically, 
to detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 (\cite{igl2010northern}). 
The approximately 1000 participants were initially mostly surveyed
about lifestyle (\cite{igl2010northern}) and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes (\cite{johansson2013identification}),
(2) the phenotypes, in this case, concentrations of certain proteins in the 
blood (\cite{enroth2014strong,enroth2015effect}).

@@

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, Ahsan and colleagues show that
SNP \verb|rs12133641| is highly correlated (p-value is $3.0^{-73}$
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
,
for $n$ = 961 individuals) with protein IL6RA (\cite{ahsan2017relative}).
The direction of the association is also concluded:
the more guanines are present at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. The amount of variance that can be explained by an association (i.e.
the \begin{math}R^2\end{math}) is rarely 100\%, which means that a trait (in
this case, the concentration of IL6RA) cannot be perfectly explained
from the genotype (in this case, SNP \verb|rs12133641|) alone. 
In this example, as much as 43\% of the variance 
can be attributed to an individuals' genotype
\iffalse
(see figure \ref{fig:ahsan2017relative_table_2_sub})
\fi
. Additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day the measurement 
was done), lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the additional variation.

@@

For computational science, it may appear to be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, 
only 18\% of 180 studies are easily reproducible \cite{stodden2018empirical}.
To some, it appears that 
the academic culture to reproduce results 
has been lost over time \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
being the exception.
One suggested way forward is to make the reproduction of 
research a minimal requirement for publication (\cite{peng2011reproducible}).

@@

A genetic epidemiologist works with sensitive data as well:
the genetic sequences of participants are private (\cite{clayton2019law}).
For research to be reproducible one needs both the code and the data
to reproduce the (hopefully) same results.
This problem is discussed in section \ref{sec:sensitive-data}.

@@

We ignore the choices of words and style of this sentence: what is
important is the content: an unpaired one-tailed T test is performed.
Taking a look at the (in this case, the programming language R (\cite{r})) code, 
we find the following line:

@@

Note that there is another side of this same coin:
correctly reporting the t-test in a reproducible ways
is known to be a problem, with only 80 out of a selected 179 papers
fully reporting the parity of t-test (i.e. paired or unpaired) 
(\cite{weissgerber2018we}).
In those unresolved 99 papers, 
the code (if published) would clarify this.
\fi

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
As humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
(\cite{baggerly2009deriving, vable2021code}),
and this example is not too far off from the
examples given in that paper.

@@

Code is rarely preserved (\cite{barnes2010publish}).
If it is, keeping it running is even more rare.
This section discusses the preservation of code for a short, medium
and long term.

@@

To preserve code for a short term, 
a code hosting website is a good first step.
A code hosting website is a website where 
its users can create dedicated pages (called 'repositories')
for a project, upload code and interact with that code.
There are multiple code hosting websites, 
with GitHub being the most popular one (\cite{cosentino2017systematic}).
For example, the \LaTeX \hspace{5mm} code of this chapter is hosted on GitHub 
as well, at \url{https://github.com/richelbilderbeek/chapter_paradata}.
The use of code hosting websites
has increased strongly (\cite{russell2018large}),
accommodates collaboration (\cite{perez2016ten})
and improves transparency (\cite{gorgolewski2016practical}),
due to its inherent computer-friendliness.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

\iffalse
An example from the author is the website (\cite{bbbqarticleissue157}),
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane}, 
as shown in Figure \ref{fig:bbbqarticleissue157}.
\fi

@@

Hosted code commonly keeps a history of file changes,
a so-called commit history,
where a commit is the description of the change, 
as well as the changes in one or more files.
For example, GitHub
uses a version control system called \verb|git| (hence its name).
In practice, this means that when a change is made to the code,
a new version is created. In the case that the change was harmful,
one can go back to an earlier version and continue again from there.
It is a general recommendation to put version control
on all human-produced data (\cite{wilson2014best}),
as well as openly working on the code from the start (\cite{jimenez2017four}).
Half of 56 Science magazine corresponding authors 
have such a version control mechanism (\cite{stodden2018empirical}).

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. 
\iffalse
Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE (\cite{etienne2020daisie})
have made to its code.
\fi
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

@@

To preserve code for a longer timespan, 
continuous integration ('CI') allows one to 
verify if code still works and, if not, to be notified on the need for an update.
Things change: from operating systems, to libraries that code uses,
to the code itself. 
Embracing those changes is recommended (\cite{beck2000extreme})
and useful code does so.

@@

Some code hosts, among others, GitHub, 
allow the researcher/user to start software scripts upon uploading a change.
This is useful to determine the reproducibility of the code: 
one such script should run the code (on trivial data) 
to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediately see which small change was the cause of this.
Using a CI service (such as GitHub Actions for GitHub), 
is known to significantly 
increase the number of bugs exposed (\cite{vasilescu2015}) and increases
the speed at which new features are added (\cite{vasilescu2015}).
Using CI opens up the possibility to informally annotate features of
a repository. 
\iffalse
For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a \LaTeX document (i.e. this paper)
could be built.
\fi

@@

For paradata to be useful, it has to be computer-friendly, yet 
'the best paradata does not necessarily look like 'data' at all for its
human users' \cite{huvila2022improving}.
There are features of code that humans find useful,
without directly being able to measure these.
In the end, code is just 'another kind of data' and should be designed 
as such, for example, by using tools to work on it (\cite{wilson2022twelve}).

@@

A first example is to use a tool to enforce a coding style 
(e.g. the Tidyverse style guide (\cite{wickham2019advanced}) for R,
or PEP 8 (\cite{van2001pep}) for Python),
as following a consistent coding style improves software quality (\cite{fang2001}).

@@

A second example is to use a tool to enforce a low cyclomatic complexity.
The cyclomatic complexity is approximately defined 
as the number of independent paths that
the code can be executed. 
\iffalse
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false,
resulting in the body of the \verb|if| statement being either
executed or ignored.
\fi
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
(\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}).

However, for code to work longer, one can put the code and its dependencies in
a container, that should be best in standing the test of time,
as the degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here a more modern (and in this context superior)
definition is used: software collapse (\cite{hinsen2019dealing}),
in which software fails due to dependencies on other 
software.
Where that paper recommends using older software foundations 
with multiple implementations (\cite{hinsen2019dealing}),
this paper recommends a modern solution that potentially has more 
longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

@@

To repeat features of useful paradata: paradata should be comprehensive, 
documented appropriately, documented in co-evolution with the data,
and the data should be friendly to computers (\cite{huvila2022improving}).
In this section, these ideal properties of paradata are applied to code.
However, code has multiple uses, which have different requirements
for being useful paradata.

Code can be used to reproduce, to replicate and to 
extend a computational experiment (\cite{benureau2018re}) and
each of these uses has different requirements for the code to be useful paradata.
In this section, the recommendations how to make paradata useful
is applied to code and compared to existing best practices in software 
development, for the intended use of the code.

@@

For code to be ideal paradata, it must be usefully 
documented (\cite{huvila2022improving}).
What is 'useful' is determined by its intended use: 
to reproduce, replicate or extend (see section \ref{sec:making-code-useful-paradata}).
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

For purposes of reproducing (same code, same data) code,
it should at least be documented how the code is run.
However, 57\% out of 56 Science papers with obtainable code (in 
total there were 180 papers) have documented their code well enough to run it 
and to be understood what it ought to do (\cite{stodden2018empirical}).

However, when an independent researcher can redo an experiment, 
it is possible (even ideal!) that no code is read at all.
Within that context, it could argued that no further documentation is needed.
However, all code in general should be documented 
'adequately' (\cite{peng2006reproducible}),
ideally writing code in such a way that it becomes 
self-explanatory (\cite{wilson2014best})
and for the remaining code to document the reasons behind it, 
its design and its purpose (\cite{wilson2014best}).

For purposes of extending a study and its code, 
documentation becomes even more important,
as the code will be read and modified.
The extent of investing time in documenting code is recommended
to be proportional to the intended reuse (\cite{pianosi2020successfully})
and there exists a clear relationship between the reuse 
of code and its documentation 
effort (\cite{cosentino2017systematic,hata2015characteristics}).

@@

For code to be ideal paradata, 
its creation and documentation need to align, 
\emph{not least because they shape each other} (\cite{huvila2022improving}).
This emphasised part of the quote resonates strongly with the idea of 
literate programming (\cite{knuth1984literate}),
in which documentation and code are developed hand-in-hand.
Literate programming is the practice of writing a document mixed with code that 
will be run.
Contemporary examples of this idea are, among others, 
vignettes (\cite{wickham2015r}) for the R programming language (\cite{r}) 
and Jupyter notebooks (\cite{wang2020assessing})
for the Julia (\cite{Julia-2017}), Python (\cite{van1995python}) and R (\cite{r})
programming languages.
rOpenSci, a community that, among others, reviews programming
code (\cite{ram2013ropensci,ram2018community}), is 
one example of where literate programming is enforced:
code must have extensive runnable examples
as part of the documentation (\cite{ropensci_2021_6619350}).
Developing code and documentation at the same time 
leads to both better code and documentation (\cite{reenskaug1989environment}).
In general, when developing software, 
it is recommended to 
to write documentation while writing software,
as well as to include many examples (\cite{lee2018ten}).

@@

Code should be distributed in standard ways (\cite{peng2006reproducible})
(see section \ref{sec:code-availability}
for the general availability of code).
The standard way to publish code is to use a code hosting website
(see subsection \ref{subsec:code-hosting}).

@@

Code must be more extensive when it is (intended to be) 
used on different data,
as then 'code must act as a teacher for future developers' (\cite{sadowski2018modern}).
Error handling is one of the mechanisms to do so.
One example is when doing a t-test:
to compare the means of two distributions, each distribution
must have at least two values
and, if not, the programming language should signal the error clearly.
In genetic epidemiology, it is common to have incomplete or missing data,
so analyses should take into account that there are not always enough
observations to do a t-test.
The analytical code should clearly signal this problem.

@@

Coding errors are extremely common 
and contribute to the reproducibility crisis in science (\cite{vable2021code}).
Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic).

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality (\cite{horgan1994,del1995correlation}). 
The non-profit organisation rOpenSci (\cite{ram2018community}),
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

@@

Having a 100\% code coverage is not uncommon,
as there is a common software development practice that 
strives for this, called test-driven development ('TDD'). 
TDD is a software development strategy in which tests are written
before new 'real' code is written.
TDD improves code quality (\cite{alkaoud2018quality,janzen2006test})
and writing comment and documentation as part of the process
is easy (\cite{shmerlin2015document}).

@@

It is considered good practice
to add a software license (\cite{jimenez2017four}),
so that it is clear that the software can be reused.
About two-thirds of 56 Science magazine corresponding authors 
that shared their code supply a software license (\cite{stodden2018empirical}).

@@

More than half of 315 scientists 
have their code rarely or never reviewed (\cite{vable2021code}),
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
(\cite{vable2021code}).
Within software development best-practices, 
code reviews are recommended as well (\cite{wilson2014best}) 
The observation that code is reviewed, makes it likelier to be correct.

@@

For those 22 (out of 180) Science magazine corresponding authors 
that supplied code and that code was estimated to be 
reproducible, one in three had the results reproduced with 
minimal or no effort (\cite{stodden2018empirical}).

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data are
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy (\cite{zhang2016review,azencott2018machine}).

Next to the code, it is the data used in an experiment 
that must be made available for an experiment 
to be called 'reproducible' (\cite{peng2006reproducible}).
When the input data are sensitive, however, and thus cannot be shared with
an independent researcher, one thus cannot reproduce the
experiment.
